Variation generation and mean fitness increase: \\ Necessity is the mother of genetic invention

### 2017-11-21 ms.tex
more UL changes

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation $f'_k$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= \big(1 - \delta C_0\big) \frac{w_0}{\bar{w}} f_0 + \beta C_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{w_k}{\bar{w}} f_k + \\
	 & \delta C_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$.
Here $\bar{w}=\sum_{k=1}^{n}{f_k w_k}$ is the population mean fitness.

@@

Similarly, if demes are arranged in a circle, for example around a lake, 
then we can denote the probability $p_k$ of migrating $k$ demes away from the parental deme (conditioned on migration which occurs with probability $C_k$)
and ${M}$ takes the form
\begin{equation}
{M} = \begin{bmatrix}
p_0 & p_1 & \ldots & p_{n-1} \\
p_{n-1} & p_0 & \ldots & p_{n-2} \\
\vdots & \vdots & \ddots & \vdots \\
p_1 & p_2 & \ldots & p_0
\end{bmatrix},
\end{equation}
where $p_k > 0$ and $\sum_{k=1}^{n}{p_k}=1$.

@@

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\hat{\bar w},
\end{equation}
where $\hat{\bar w}$ is the stable population mean fitness in the absence of modifier allele $M$.
Note that, in general, $\hat{\bar w}$ is not an eigenvalue of $\cl_{ex}$.

\begin{theorem}[Evolution of increased genetic variation]
Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.
\end{theorem}

@@

Eq.~\ref{mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \emph{\nameref{sec:mut_model_2}} section).
Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.

>>>>

Let the frequency of $A_k$ in the present generation be $f_k$ with $f_k \ge 0$ and $\sum_{k=1}^{n}{f_k}=1$.
Then after selection and mutation $f'_k$ in the next generation is given by
\begin{equation}
\begin{aligned}
f_0' &= \big(1 - \delta C_0\big) \frac{w_0}{\bar{w}} f_0 + \beta C_{1} \frac{w_{1}}{\bar{w}} f_{1}, \\
f_n' &= \big(1 - \beta C_n\big) \frac{w_n}{\bar{w}} f_n + \delta C_{n-1} \frac{w_{n-1}}{\bar{w}} f_{n-1}, \\
f_k' &= \big(1 - C_k(\delta+\beta)\big) \frac{w_k}{\bar{w}} f_k + \\
	 & \delta C_{k-1} \frac{w_{k-1}}{\bar{w}} f_{k-1} + 
	 \beta C_{k+1} \frac{w_{k+1}}{\bar{w}} f_{k+1},
\end{aligned}
\label{eq:mutation_model_2}
\end{equation}
for $k=1,2,\ldots, n-1$.
Here $\bar{w}=\sum_{k=0}^{n}{f_k w_k}$ is the population mean fitness.

@@

Similarly, if demes are arranged in a circle, for example around a lake, 
then we can denote the probability $p_k$ of migrating $k$ demes away from the parental deme (conditioned on migration which occurs with probability $C_k$)
and ${M}$ takes the form
\begin{equation}
{M} = \begin{bmatrix}
p_0 & p_1 & \ldots & p_{n-1} \\
p_{n-1} & p_0 & \ldots & p_{n-2} \\
\vdots & \vdots & \ddots & \vdots \\
p_1 & p_2 & \ldots & p_0
\end{bmatrix},
\end{equation}
where $p_k > 0$ and $\sum_{k=0}^{n-1}{p_k}=1$.

@@

$\cl_{ex}$ can be written as
\begin{equation}\label{eq:Lex}
\cl_{ex} = (I - \tilde{C} + M \tilde{C}) D /\hat{\bar w},
\end{equation}
where $\hat{\bar w}$ is the stable population mean fitness in the absence of modifier allele $M$.
Note that, in general, $\hat{\bar w}$ is not an eigenvalue of $\cl_{ex}$.
\medskip

\noindent{\bf Result 2. Evolution of increased genetic variation.}
{\sl Let $\lambda_1$ be the the leading eigenvalue of the external stability matrix $\cl_{ex}$.
If the transition rates induced by the modifier alleles $m$ and $M$ are equal, i.e., $\tilde{C}_k=C_k$ for all $k$, then the following applies
\begin{equation}
\lambda_1=1,
\end{equation}
and
\begin{equation}
sign \frac{\partial \lambda_1}{\partial \tilde{C}_k}\bigg\rvert_{\tilde{C}_k = C_k} = 
sign(\hat{\bar w} - w_k).
\end{equation}

Therefore, an initially rare modifier allele $M$ with transition rates slightly different from the resident allele $m$ can successfully invade the population ($\lambda_1>1$) if $M$ increases the probability of transition from types with below-average fitness, thereby increasing the stable population mean fitness.}

@@

Eq.~\ref{mfp} is a generalization of a result of~\citet[Eq.~4]{Ram2012}.
\citeauthor{Ram2012} modeled the accumulation of mutant alleles in a population evolving at the mutation-selection balance (see \nameref{sec:mut_model_2} section).
Using Eq.~\ref{eq:Caswells_formula} in \nameref{sec:AppA} and a recursion on the ratios of the reproductive values \cite[see][eqs.~A5-6]{Ram2012}, they concluded that at the mutation-selection balance, if individuals with below-average fitness ($w_k < \hat{\bar w}$) increase their mutation rate, then the population mean fitness will increase -- a result generalized by our \emph{mean fitness principle} in Eqs.~\ref{mfp_sign} and~\ref{mfp_sign_threshold}.
